Universal Characterization of Physical Variability by \\Shannon's $-\sum_ip_i\,\textrm{ln}\,p_i$ Formula

### 2019-10-22 Articles/2019-ShannonEntropy/Source/ShannonEntropyIsVariability.tex
Adding Boltzmann entropy and bibliography

The first question is straightforward to answer. The macroscopic variables of a system (e.g. volume, pressure, energy) do not give a full description of its parts. If we fix a macrostate and then measure the state of all its microscopic constituents, we will not get the same answer every time. This true in the case of non equilibrium (e.g. we suddenly compress a gas pushing a piston) and of equilibrium (e.g. we give enough time to equilibrate). The choice of a macrostate, then, identifies a set of possible microstates that are compatible with it. It is then valid to ask the question: what is the variability of microstates one can find within a given macrostate? This is answered by the Shannon variability, which gives us the number of bits required to identify a microstate given the macrostate. Since the macrostates fully define the distribution of possible microstates, this distribution does not depend on the particular way the macrostate came to pass. It is path independent: it is a state variable. In short, as a macrostate is a distribution over microstates, it will admit as a state variable the Shannon variability, regardless of whether we know the actual distribution.

Note, though, that the thermodyanmic entropy is defined differently. It is defined over a succession of equilibrium states by noting that the integral of the heat exchanged divided by the temperature is path independent. One then chooses a particular state $A$, assigns it an arbitrary value $S_A$ and the entropy for all other states is given by the integral $S(B) = S_A + \int_A^B \frac{dQ}{T}$. At this point, it is not apparent why these two quantities defined independently would be related. What we can say, though, is that all quantities that increase over irreversible processes, are maximized at equilibrium and that are additive for indipendent systems are precisely the quantities that are linear functions of entropy.[CITE AXIOMATIC WORK] A linear function of entropy, though, simply means a redefinition of $S_A$ and a change of units for either $dQ$ or $T$. Therefore if we can show that the Shannon variability possesses those three properties than it is equivalent to the entropy up to a change in units, which would answer our second question.

Suppose we have an irreversible process, which means we can start from many out-of-equilibrium states which end into a single equilibrium state. During the whole process, some of the macro variables are going to stay constant as they characterize the constraint the system is subjected during the evolution. For example, if the walls are fixed, the gas will have a set volume; if the walls are insulated, then the average energy will be conserved.\footnote{It can also happen that the irreversible process does have a single possible equilibrium, as in the case of spontaneous symmetry breaking. New variables will identify the equilibrium which we can think of new constraints appeared once the initially symmetry is broken.} As different out-of-equilibrium macrostates merge during the evolution, they merge the possible initial conditions that lead to the same macrostate. The variability of microstates within the initial out-of-equilibrium macrostate, then, must be smaller than the variability of the microstates within the final equilibrium state. The amount of information needed to identify a microstate at equilibrium allows us to identify the initial microstate across multiple out-of-equilibrium states and therefore is greater than the information needed to identity a microstate in an out-of-equilibrium macrostate. Irreversible processes increase the Shannon variability within the macrostate, which corresponds to a greater uniformity of the macrostate itself since uniform distributions are the ones that maximize the variability.

@@

Now suppose we have two independent systems, like two volumes of gasses. As they are independent, picking the microstate of one system does not constrain the microstate of the other. That is, the distribution over microstates are statistically independent. This means that the statistical distribution is the product of the join distributions and, in this case, the Shannon entropy is additive.

@@

The first is the role of the Boltzmann constant $k_B$, which is a conversion factor between the Shannon variability. $H = - \sum p_i \log(p_i)$ and the Gibbs entropy $S_G = - k_B p_i \sum \log(p_i)$. This conversion constant should not surprise us, since we only established a linear relationship between variability and entropy. The question is whether this conversion constant itself has a physical meaning, or it is also simply due to a choice of units. The standard relationship between temperature, internal energy and entropy is $T = \frac{\partial U}{\partial S}$. Historically, temperature and internal energy are the quantities that were defined first. If we measure temperature in Kelvin and energy in Joules, the above relationship forces us to measure entropy in Joules/Kelvin. If we changed units, $k_B$ would change accordingly.

However, this is not the only way to define the relationships between those quantities. Another way is to have $\beta = \frac{1}{k_B T} = \frac{\partial S}{\partial U}$. In this setting, $U$ is measured in Joules, $S$ in nats (which are dimensionless), $\beta$ is measured in inverse Joules. Temperature is still measured in Kelvin, but we do not regard it as a primary object. We find this approach, sometimes called entropy first, preferable for a number of reasons, especially in statistical mechanics. First, entropy is invariant under coordinate transformations and therefore it is better to keep it a pure number and not mix it with units of energy, which is not invariant under boost.\footnote{This is essential if one ever hopes to put statistical mechanics in a relativistic setting.} Second, the maximization of entropy at equilibrium guaranties us that $S(U)$ is a single valued function (i.e. the maximum given the value of $U$) and $\frac{\partial S}{\partial U}$ is therefore well defined. On the other hand, $U(S)$ is not in general single valued (e.g. in the Ising model). This means that temperature can be infinite where the curve inverts, where small changes of energy do not change the entropy, and can be negative in the other region. That is, the infinitely cold system is temperature zero, which cannot be achieved, as it warms up temperature rises until it becomes infinite, and as the system becomes even hotter the temperature is negative and increase toward zero, which would correspond to an infinitely hot system. On the other hand, $\beta$ is always finite and cannot be infinite. An infinitely cold system, then would correspond to positive infinite $\beta$, as the system warms up beta decreases, passes from zero, becomes negative and an infinitely hot system corresponds to minus infinite $\beta$: this is a much better behavior for a physically meaningful quantity. Third, when setting up entropy maximization, $\beta$ is quite naturally the Lagrange multiplier associated with energy conservation, therefore mathematically it has a much clearer role. Lastly, one can proceed in both thermodynamic and statistical mechanics with just $\beta$, and convert to temperature only when we need a value on the Kelvin scale. For these reasons, it seems to us $\beta$ plays the more fundamental role and $k_B$ is the conversion coefficient to go from $\beta$ to temperature measured in Kelvin. Adding and removing the Boltzmann constant from the entropy expression, then, is equivalent to shift from an energy/temperature first approach to an entropy/$\beta$ first approach and should be not seen as physically significant. It is really just fixing a unit.

@@

We want to stress that a distribution on $\mu$-space can only either describe a single microstate where $N$ is very large and therefore the distribution can be assumed to be continuous, or a probability distribution for a $N$ identically and independently prepared particles. It one wanted to describe a macrostate, one would have to work with a probability distribution (i.e. the weights represents probabilities) over the space of particle distributions (i.e. the weights represent fractions of total particles) over $\mu$-space or one would obtain erroneous results.\footnote{This happens, for example, in [Jaynes] where he creates the distribution over $\mu$-space by averaging over all microstates. This essentially blends together the probabilities and particle counts losing most of the actual statistical information. Naturally, the entropy thus computed is wrong and found to not work.}

@@

This means that we always have to consider distributions over the full phase-space with the appropriate conjugate coordinates if we want to use $-\rho \log \rho$. If we use distributions over position and velocities or over momentum only, the expression would change for those particular variables in those coordinate systems. This can be, again, a source of confusion.

These technical details are important to use the Shannon variability correctly, but they do not alter the conceptual point. The Shannon variability, the thermodynamic entropy and the Gibbs entropy, though describing different things, are conceptually related. The variability of the state of the parts within the state of the whole is what the Shannon variability quantifies if applied to the macrostates of thermodynamics and the microstates of statistical mechanics. This variability increases during irreversible evolution, more microstates converge to the same macrostate until equilibrium is reached, where the variability is maximized. The variability of one system does not affect the variability of an independent one and therefore it simply sums. This is not a coincidence: this is what thermodynamic entropy ultimately is about.

@@

\bibliographystyle{unsrt}  
\bibliography{references}

>>>>

The first question is straightforward to answer. The macroscopic variables of a system (e.g. volume, pressure, energy) do not give a full description of its parts. If we fix a macrostate and then measure the state of all its microscopic constituents, we will not get the same answer every time. This is true in the case of non-equilibrium (e.g. we suddenly compress a gas pushing a piston) and of equilibrium (e.g. we give enough time to equilibrate). The choice of a macrostate, then, identifies a set of possible microstates that are compatible with it. It is then valid to ask the question: what is the variability of microstates one can find within a given macrostate? This is answered by the Shannon variability, which gives us the number of bits required to identify a microstate given the macrostate. Since the macrostates fully define the distribution of possible microstates, this distribution does not depend on the particular way the macrostate came to pass. It is path independent: it is a state variable. In short, as a macrostate is a distribution over microstates, it will admit as a state variable the Shannon variability, regardless of whether we know the actual distribution.

Note, though, that the thermodyanmic entropy is defined differently. It is defined over a succession of equilibrium states by noting that the integral of the heat exchanged divided by the temperature is path independent. One then chooses a particular state $A$, assigns it an arbitrary value $S_A$ and the entropy for all other states is given by the integral $S(B) = S_A + \int_A^B \frac{dQ}{T}$. At this point, it is not apparent why these two quantities defined independently would be related. What we can say, though, is that all quantities that increase over irreversible processes, are maximized at equilibrium and that are additive for indipendent systems are precisely the quantities that are linear functions of entropy (see \cite{Giles}). A linear function of entropy, though, simply means a redefinition of $S_A$ and a change of units for either $dQ$ or $T$. Therefore if we can show that the Shannon variability possesses those three properties than it is equivalent to the entropy up to a change in units, which would answer our second question.

Suppose we have an irreversible process, which means we can start from many out-of-equilibrium states which end into a single equilibrium state. During the whole process, some of the macro variables are going to stay constant as they characterize the constraint the system is subjected during the evolution and will also identify the final equilibrium. For example, if the walls are fixed and insulated, the gas will have a set volume and average energy, which will find a unique equilibrium with a given temperature and pressure.\footnote{It can also happen that the irreversible process does not have a single possible equilibrium, as in the case of spontaneous symmetry breaking. New variables will identify the equilibrium which we can think of new constraints, appeared once the initially symmetry is broken, thus reducing to the previous case.} As different out-of-equilibrium macrostates merge during the evolution, they merge the possible initial conditions that lead to the same macrostate. The variability of microstates within the initial out-of-equilibrium macrostate, then, must be smaller than the variability of the microstates within the final equilibrium state. The number of bits needed to identify a microstate at equilibrium allows us to identify the initial microstate across multiple out-of-equilibrium states and therefore is greater than the number of bits needed to identity a microstate in an out-of-equilibrium macrostate. Irreversible processes increase the Shannon variability within the macrostate, which corresponds to a greater uniformity of the macrostate itself since uniform distributions are the ones that maximize the variability.

@@

Now suppose we have two independent systems, like two volumes of gasses. As they are independent, picking the microstate of one system does not constrain the microstate of the other. That is, the distribution over microstates are statistically independent. This means that the joint distribution is the product of the marginal distributions and, in this case, the Shannon entropy is additive.

@@

The first is the role of the Boltzmann constant $k_B$, which is a conversion factor between the Shannon variability $H = - \sum p_i \log(p_i)$ and the Gibbs entropy $S_G = - k_B p_i \sum \log(p_i)$. This conversion constant should not surprise us, since we only established a linear relationship between variability and entropy. The question is whether this conversion constant itself has a physical meaning, or it is also simply due to a choice of units. The standard relationship between temperature, internal energy and entropy is $T = \frac{\partial U}{\partial S}$. Historically, temperature and internal energy are the quantities that were defined first. If we measure temperature in Kelvin and energy in Joules, the above relationship forces us to measure entropy in Joules/Kelvin. If we changed units, $k_B$ would change accordingly.

However, this is not the only way to define the relationships between those quantities. Another way is to have $\beta = \frac{1}{k_B T} = \frac{\partial S}{\partial U}$. In this setting, $U$ is measured in Joules, $S$ in nats (which are dimensionless), $\beta$ is measured in inverse Joules. Temperature is still measured in Kelvin, but we do not regard it as a primary object. We find this approach, sometimes called entropy first, preferable for a number of reasons, especially in statistical mechanics. First, entropy is invariant under coordinate transformations and therefore it is better to keep it a pure number and not mix it with units of energy, which is not invariant under boost.\footnote{This is essential if one ever hopes to put statistical mechanics in a relativistic setting.} Second, the maximization of entropy at equilibrium guaranties us that $S(U)$ is a single valued function (i.e. the maximum given the value of $U$) and $\frac{\partial S}{\partial U}$ is therefore well defined. On the other hand, $U(S)$ is not in general single valued (e.g. in the Ising model). This means that temperature can be infinite where the curve inverts, where small changes of energy do not change the entropy, and can be negative in the other region. That is, the infinitely cold system is temperature zero, which cannot be achieved, as it warms up temperature rises until it becomes infinite, and as the system becomes even hotter the temperature is negative and increase toward zero, which would correspond to an infinitely hot system. On the other hand, $\beta$ is always finite and cannot be infinite. An infinitely cold system, then would correspond to positive infinite $\beta$, as the system warms up beta decreases, passes from zero, becomes negative and an infinitely hot system corresponds to minus infinite $\beta$: this is a much better behavior for a physically meaningful quantity. Third, when setting up entropy maximization, $\beta$ is quite naturally the Lagrange multiplier associated with energy conservation, therefore mathematically it has a much clearer role. Lastly, one can proceed in both thermodynamic and statistical mechanics with just $\beta$, and convert to temperature only when a value on the Kelvin scale is needed. For these reasons, it seems to us $\beta$ plays the more fundamental role and $k_B$ is the conversion coefficient to go from $\beta$ to temperature measured in Kelvin. Adding and removing the Boltzmann constant from the entropy expression, then, is equivalent to shift from an energy/temperature first approach to an entropy/$\beta$ first approach and should be not seen as physically significant. It is really just fixing a unit.

@@

We want to stress that a distribution on $\mu$-space can only either describe a single microstate where $N$ is very large and therefore the distribution can be assumed to be continuous, or a probability distribution for $N$ identically and independently prepared particles. A macrostate cannot be described by a single distribution on $\mu$-space as this cannot capture correlations. One would need to work with a probability distribution (i.e. the weights represents probabilities) over the space of particle distributions (i.e. the weights represent fractions of total particles) over $\mu$-space or one would obtain erroneous results.\footnote{Unfortunately this happens, for example, in [Jaynes] where he creates the distribution over $\mu$-space by averaging over all microstates. This essentially blends together the probabilities and particle counts losing most of the actual statistical information. Naturally, the entropy thus computed is wrong and found to not work.}

@@

This means that we always have to consider distributions over the full phase-space with the appropriate conjugate coordinates if we want to use $-\rho \log \rho$. If we use distributions over position and velocities or over momentum only, the expression would change for those particular variables in those coordinate systems. This can be, again, a source of confusion.\cite{Dunkel}

The last issue is the relationship with the Boltzman entropy $S_B= k_B \log W$. This is actually captured by another standard way to derive the Shannon variability. Suppose we have a sequence $N$ elements, say twenty balls. There are $I$ possible types of elements, say three possible colors: red, green and blue. Suppose $N_i$ are the elements of each possible types, which means $\sum_{i \in I} N_i = N$. Then, given a set of $N_i$, we have $W(N_i) = \frac{N!}{\prod_{i \in I} N_i}$ possible ways to realize that sequence which corresponds to all possible permutations. If all permutations are equally likely, then $\log W$ represents the number of bits needed to identify one of the sequences. Since there are $N$ elements in the sequence, $\frac{1}{N} \log W$ represents the average number of bits needed to identify one element. We now take the limit for large $N$, that is we ask what is the average number of bits needed to identify one element from an infinite sequence given by the weights $p_i = \frac{N_i}{N}$. We find $\frac{1}{N} \log \frac{N!}{\prod_{i \in I} N_i} \approx - \sum p_i \log p_i$. In other words, the Boltzman entropy reduces to the Shannon entropy when counting permutations within large distributions.

These technical details are important to understand the fine details of the links between the different entropies, they do not alter, and should not distract us, from the main conceptual point. The Shannon variability, the thermodynamic entropy, the Boltzmann entropy and the Gibbs entropy, though describing different things, are conceptually related. The variability of the state of the parts within the state of the whole is what the Shannon variability quantifies if applied to the macrostates of thermodynamics and the microstates of statistical mechanics. This variability increases during irreversible evolution, more microstates converge to the same macrostate until equilibrium is reached, where the variability is maximized. The variability of one system does not affect the variability of an independent one and therefore it simply sums. This is not a coincidence: this is what thermodynamic entropy ultimately is about.

@@

\bibliographystyle{unsrt}  
\bibliography{bibliography}
